---
title: "Exploratory Data Analysis for the 2019 Ridgecrest Sequence"
author: "Sachit Ramakrishnan"
bibliography: 01_eda_ridgecrest.bib
format: html
editor: visual
---

# Introduction

## What is an earthquake catalogue?

An **earthquake catalogue** is a table with one row per detected earthquake. For each event, a **seismic network** estimates *when* it happened (time of origin), *where* it happened (the longitude and latitude of the epicentre, and the depth below the surface), and *how large* it was (its magnitude). These are all estimates derived from the ground motion recorded at many stations, so they carry a level of uncertainty and occasional processing quirks. The purpose of this notebook is to understand the features and limitations of the data *before* modelling.

## What is the ETAS model?

Earthquakes are not independent: a large earthquake triggers a sequence of (usually smaller) events, whose rate decays with time. The **Epidemic-Type Aftershock Sequence (ETAS)** [@ogata1988] model captures this by assuming the rate of earthquakes at any moment is a **background** rate (assumed constant here) plus a **triggering** contribution from every past event. This project fits a **temporal** ETAS model (time only, ignoring space) to the **2019 Ridgecrest sequence**, and compares different parametric forms for the triggering component. For that, we first need clean data, a completeness threshold, and a defined sequence. These will be provided by the exploratory analysis.

## What is the purpose of this notebook?

The key goal with Exploratory Data Analysis (EDA) is to summarise key features of the data before building models. In our case, this means to:

1.  **understand the data and find artefacts** - features caused by how the data were recorded/processed rather than by real earthquakes;
2.  **extract the model's inputs** - the **magnitude of completeness $M_c$** (the smallest magnitude above which almost all events are recorded) and the Gutenberg-Richter **$b$-value** (which determines the magnitude of events);
3.  **define the sequence** - choose the spatiotemporal window that isolates the Ridgecrest sequence;
4.  **assess limitations and validity of assumptions** presented by the temporal ETAS model; motivate comparison of triggering kernels.

The workflow largely follows the exploratory analysis developed for the `ETAS.inlabru` package [@naylor2023], available at <https://github.com/edinburgh-seismicity-hub/exploratory-seismicity-analysis/tree/main>. Here, that analysis is applied to a new data set and extended for our temporal-kernel focus.

## The data

This notebook analyses `2026-07-17_scedc-scsn_socal_2016-2025_raw.txt`: the **Southern California Seismic Network (SCSN) [@scsn1926] earthquake catalogue** (2016-2025), obtained from the **Southern California Earthquake Data Center (SCEDC) [@scedc2013]**. **2016** was chosen as the start year to ensure a consistent processing regime [@scedc_changehistory], and **2025** as the end year as the last complete calendar year of processed data [@scedc_cataloguestatus]. We impose no restrictions on magnitude, depth, or event type. Additionally, the latitude/longitude search bounds were left unrestricted, with the geographic restriction imposed by the local geographic-type filter. This raw catalogue is later filtered to earthquakes and subset to the Ridgecrest sequence during analysis. 

**The data's source and full documentation are detailed in `data/raw/metadata.txt`.**

# Setup

```{r setup}
################################################################################
# Chunk options 
################################################################################
knitr::opts_chunk$set(
  echo      = TRUE,      
  message   = FALSE,     
  warning   = FALSE,     
  fig.align = "center",
  fig.width = 7, fig.height = 4.5, dpi = 150
)

################################################################################
# Packages 
################################################################################
library(dplyr)
library(geosphere)
library(ggplot2)
library(ggrepel)
library(here)
library(lubridate)
library(maps)
library(sf)
library(stats)
library(tidyr)
library(zoo) 
################################################################################
# Project functions (see /src) 
################################################################################
source(here("src", "read_catalogue.R"))
source(here("src", "eda_helpers.R"))
source(here("src", "bvalue.R"))

################################################################################
# Global plot theme 
################################################################################
theme_set(theme_bw())
################################################################################
# Reusable objects
################################################################################
# Base-map polygons: computed once here, reused by every map
world     <- map_data("world")
us_states <- map_data("state")

base_map <- list(
    # Land fill: light-grey world polygons, no outline
    geom_polygon(data = world, aes(long, lat, group = group),
                 fill = "grey97", colour = NA),
    # Internal US state borders: thin and light
    geom_polygon(data = us_states, aes(long, lat, group = group),
                 fill = NA, colour = "grey75", linewidth = 0.25),
    # National borders: darker, on top of the state lines
    geom_polygon(data = world, aes(long, lat, group = group),
                 fill = NA, colour = "grey30", linewidth = 0.5)
  )

# Shared earthquake / quarry-blast colours and labels
et_cols <- c(eq = "steelblue", qb = "firebrick")
et_labs <- c(eq = "Earthquakes", qb = "Quarry blasts")

# Magnitude bin width - used to discretise magnitudes when building
# frequency-magnitude distribution
dm <- 0.1
```

**`read_catalogue()`** reads the raw SCEDC file, discards header and footer lines, assigns appropriate names for the columns, and creates a new date-time field (among a few other convenient columns).

```{r load}
# Load and parse data
cat_path <- here("data","raw", "2026-07-23_scedc-scsn_socal_2016-2025_raw.txt")
cat_raw_all<- read_catalogue(cat_path)
```

```{r important-observations}
# Ridgecrest mainshock (M7.1) and foreshock (M6.4), in UTC.
mainshock <- cat_raw_all%>%filter(mag == 7.1, date == "2019/07/06")
mainshock_time <- mainshock$datetime
foreshock <- cat_raw_all%>%filter(mag==6.4, date == "2019/07/04")
foreshock_time <- foreshock$datetime
mainshock_lon  <- mainshock$lon
mainshock_lat  <-   mainshock$lat
foreshock_lon  <- foreshock$lon
foreshock_lat  <-   foreshock$lat

events <- data.frame(
  lon   = c(mainshock_lon, foreshock_lon),
  lat   = c(mainshock_lat, foreshock_lat),
  time  = c(mainshock_time, foreshock_time),
  type  = c("Mainshock", "Foreshock")
)
```
## Geography type

```{r map-all}
cat_raw_all %>%
  ggplot() +
  base_map +
  geom_point(aes(lon, lat, colour = gt), size = 1, alpha = 0.5) +
  geom_point(data = events, aes(lon, lat, shape = type), size = 3, stroke = 1.2) +
  scale_shape_manual(values = c(Mainshock = 8, Foreshock = 5), name = "Key") +
  scale_colour_manual(
    values = c(l = "grey55", r = "blue", t = "purple"),
    labels = c(l = "Local", r = "Regional", t = "Teleseismic"),
    name   = "Geographic type"
  ) +
  coord_quickmap() +
  labs(title = "Events by geographic type", x = "Longitude", y = "Latitude")
```

We choose to retain events of local geographic type (`gt=l`). The reasoning behind this is physically motivated. The $M7.1$ Ridgecrest produced ~46 km of surface rupture [@barnhart], and aftershocks are expected to occur within 1-2 fault lengths of the rupture [@usgs_aftershocks]; consistently, Gardner and Knopoff [@gardner1974sequence] specify an aftershock distance window of ~70km for an M$7.0$ South California earthquake, aftershock. Because these empirical estimates are derived from observed sequences, they already encompass higher-generation aftershock, which are triggered locally and remain spatially confined. To test for contamination conservatively, we consider an ellipse aligned with the $\sim 322^{\circ}$ rupture strike [@barnhart], with semi=major axis three times and semi-minor axis two times the fault length - exceeding both estimates above. No regional or teleseismic events fall within this conservative zone, validating their removal from analysis. 

```{r map-aftershock}
# Ellipse parameters
strike_deg <- 322          # fault strike, NW-SE (Ridgecrest M7.1 ~ 322°)
a <- 46 * 3 * 1000         # semi-major axis along strike
b <- 46 * 2 * 1000         # semi-minor axis across strike (e.g. 18 km)

# Build oriented ellipse
t       <- seq(0, 2*pi, length.out = 361)
along   <- a * cos(t)                       # component along strike
across  <- b * sin(t)                       # component across strike
dist    <- sqrt(along^2 + across^2)         # distance from centre
bearing <- (strike_deg + atan2(across, along) * 180/pi) %% 360   # bearing from N

ellipse <- as.data.frame(
  destPoint(p = c(mainshock_lon, mainshock_lat), b = bearing, d = dist)
)  

# Plot
cat_raw_all %>%
  filter(gt != "l")%>%
  ggplot() +
  base_map +
  geom_point(aes(lon, lat,col=gt), alpha = 0.4) +
  geom_path(data = ellipse, aes(lon, lat), colour = "red", linewidth = 0.6) +
  geom_point(data = events, aes(lon, lat, shape = type), size = 3, stroke = 1.2) + 
  scale_shape_manual(values = c(Mainshock = 8, Foreshock = 5), name = "Key") +
  scale_colour_manual(
    values = c(r = "blue", t = "purple"),
    labels = c(r = "Regional", t = "Teleseismic"),
    name   = "Non-local events"
  ) +
  coord_quickmap(xlim = mainshock_lon + c(-2,2),
                 ylim = mainshock_lat + c(-2,2)) +
  labs(title = "Ridgecrest sequence and fault-aligned aftershock zone",
       subtitle = "Ellipse aligned NW-SE along the M7.1 rupture",
       x = "Longitude", y = "Latitude")
```



```{r local}
# Retain only local events
cat_raw <- cat_raw_all%>%filter(gt == "l")
```

## Summary of local events

```{r map}
# Initial map: whole world, to reveal any location oddities
ggplot() +
  base_map +
  geom_point(data = cat_raw, aes(lon, lat),
             colour = "firebrick", size = 0.4, alpha = 0.5) +
  geom_point(data = events, aes(lon, lat, shape = type), size = 3, stroke = 1.2) +
  scale_shape_manual(values = c(Mainshock = 8, Foreshock = 5), name = "Key")+
  coord_quickmap() +
  labs(title = "'Local' events, world view",
       subtitle = "Note the cluster at 0\u00B0N, 0\u00B0E ('Null Island')",
       x = "Longitude", y = "Latitude")
```

Plotting the raw catalogue on a world map immediately reveals a problem: a cluster of events that should be confined to the SCSN region instead are at $(0^{\circ}N ,0^{\circ}E)$ - a point in the Atlantic Ocean known as *Null Island*.

```{r null-island}
# Extract events with 'Null Island" location
null_island <- cat_raw %>% filter(lat == 0, lon == 0)
null_island
```

Zero picked phases (`nph`) for all observations indicates that the network detected an event but had too few phases to locate it, so the catalogue stored a placeholder coordinate of $(0,0)$. They are therefore data artefacts rather than genuine seismicity and are removed, although the precise cause of each cannot be determined from the catalogue alone.

```{r remove-null-island}
# Remove 'Null Island' events
cat_clean <- cat_raw %>% filter(!(lat == 0 & lon == 0))
```


```{r summary}
# Cleaned-catalogue summary 
cat("Events:", nrow(cat_clean),
    "| Window:",format(min(cat_clean$datetime)), "to",
    format(max(cat_clean$datetime)), "\n" )
```

```{r map-local}
# Map of the cleaned catalogue, cropped to its spatial extent
spatial_window <- cat_clean%>%
  summarise(lat_min = min(lat), lat_max = max(lat),
            lon_min = min(lon), lon_max = max(lon))

ggplot() +
  base_map +
  geom_point(data = cat_clean, aes(lon, lat),
             colour = "firebrick", size = 0.3, alpha = 0.3) +
  geom_point(data = events, aes(lon, lat, shape = type), size = 3, stroke = 1.2) +
  scale_shape_manual(values = c(Mainshock = 8, Foreshock = 5), name = "Key")+
  coord_quickmap(
            xlim = c(spatial_window$lon_min - 0.5, spatial_window$lon_max + 0.5),
            ylim = c(spatial_window$lat_min - 0.5, spatial_window$lat_max + 0.5)) +
  labs(title = "Local events", x = "Longitude", y = "Latitude")
```

With the placeholder records removed, the catalogue behaves as expected: the great majority of events lie in southern California, with a minority extending south into Baja California (Mexico) and the Gulf of California, and a scattering of events offshore and towards the Nevada border. 

## Event type

The raw download contains more than just earthquakes detected by the SCSN. 
The *event type* (`et`) column records *earthquakes* (`eq`), in addition to other sources of seismic activity. This field is determined by analysts, based on factors such as event time, location and waveform characterisation [@allmann2008].
We retain the whole catalogue for inspection, and filter to earthquakes only once we have confirmed that the event types are separable.

### What is in the catalogue?

```{r type-counts}
# Count summaries for event type and geography type
nrow(cat_clean)
cat_clean %>% count(et, sort=TRUE)
cat_clean %>% count(gt, sort=TRUE)
```
The catalogue is overwhelmingly comprised of earthquakes: $239,140$ events ($97.1 \%$). The largest non-earthquake category is **quarry blasts** ($7,173; 2.91\%$), while other man-made events are negligible ($25$ *explosions of unknown cause*, $3$ *sonic booms*).
As expected, all events are **local** to the SCSN.

### Sonic booms and explosions of unknown cause

```{r sn-ex}
# Summary of sonic boom and unknown explosion events
cat_clean %>%
  filter(et %in% c("ex", "sn")) %>%
  group_by(et) %>%
  summarise(n= n(),
            mag_range = paste(min(mag), "to", max(mag)),
            depth_med = median(depth),
            first     = min(datetime),
            last      = max(datetime),
            .groups   = "drop")

# Extracting the most significant events
cat_clean %>%
  filter(et %in% c("ex", "sn")) %>%
  filter(mag>=1.5)
```           

We remove all `ex` ($ \sim 0.013 \%$) and `sn` ($ \sim 0.001 \%$) events, since neither is tectonic seismicity. Due to their negligible numbers, the decision is insensitive to label reliability: even if every one were a misclassified earthquake, removing them would not materially affect model fit. The largest event in this subset has magnitude $1.71$, so these are **small** events, and their removal is insensitive to the completeness threshold we determine later. If $M_c$ exceeds $\sim 1.71$, all events are excluded by the completeness cut, and even under a conservative $M_c$ of $1.5$, only a single event would be retained. We therefore remove them.

## Earthquakes vs quarry blasts

The ETAS framework we use rests on a generative assumption: every event in the catalogue is either **background** seismicity (driven by underlying tectonic processes) or **triggered** by a preceding earthquake - there is no third category. Quarry blasts fit neither: they are anthropogenic, externally scheduled events with no tectonic cause. Retaining them would therefore bias model parameters, as the model would be forced to attribute each blast to background or triggering. 
Before removing them, however, we verify that `qb` and `eq` events form genuinely separable populations: we compare their characteristics in time, size, depth and space. This is important because we are acting on the catalogue's analyst-assigned labels: only if blasts form a distinct, identifiable group can we trust that removing `qb`-labelled events strips out genuine blasts rather than misclassified earthquakes. Should the two be indistinguishable, the labels would be unreliable, and we would retain the events rather than risk excluding earthquakes non-randomly.
Individual mislabelling (earthquakes tagged as blasts or vice versa) cannot be excluded, and modern machine-learning approaches would likely classify some events differently [@zhu2024]. We nonetheless adopt the catalogue labels and note residual mislabelling as a limitation.

### Time

```{r et-df}
# Create data frame with only `qb` and `eq` events
eq_qb <- cat_clean %>% 
  filter(et %in% c("eq", "qb"))
```

```{r et-ts}
# Monthly count of each event type over time
eq_qb%>%
  count(et, month_start) %>%
  ggplot(aes(month_start, n, colour = et)) +
  geom_line(linewidth = 0.5)+
  scale_colour_manual(values = et_cols,
                    labels = et_labs)+
  facet_wrap(~ et, ncol = 1, scales = "free_y",
             labeller = as_labeller(et_labs)) +   
  labs(title = "Monthly event counts: earthquakes vs quarry blasts (2016-2025)",
       x = "Month", y = "Events per month", colour = "Type") 
```

The two series differ in both **scale** and **character**. The panels have independent y-axes: on a shared axis the blasts would be invisible. Quarry blasts run at roughly 10-100 per month, while earthquakes range from about 1,000-2,000 in quiet months to tens of thousands (~30,000) at the height of the Ridgecrest sequence.
Their temporal character differs just as sharply. Quarry blasts are generally **steady** across the window - a consistent, human-scheduled background process. The time series displays a slight downwards trend, in addition to a clear seasonality, which will be explored in later plots. Earthquakes, on the other hand, are **episodic**: a flat background until July 2019, then a sudden spike as the mainshock and its aftershocks arrive, decaying back towards background over the following two to three years. This highlights that blasts, if retained, would likely upwardly bias the background parameter.
```{r qb-seasonality}
# Plotting monthly quarry blasts for each year, with 2019 highlighted in 
# comparison to an average year (not including 2019)
qb_month <- cat_clean %>%
  filter(et == "qb") %>%
  count(year, month)

avg_year <- qb_month %>%
  filter(year != 2019) %>%
  group_by(month) %>%
  summarise(n = mean(n), .groups = "drop")

year_ends <- qb_month %>%
  group_by(year) %>% slice_max(as.integer(month), n = 1) %>% ungroup()

ggplot(qb_month, aes(month, n, group = year)) +
  geom_line(colour = "grey80", alpha = 0.8) +
  geom_line(data = avg_year, aes(group = 1), colour = "black", linewidth = 1.2) +
  geom_line(data = subset(qb_month, year == 2019), colour = "red", linewidth = 1.2) +
  geom_text_repel(data = year_ends, aes(label = year),
                  size = 3, hjust = 0, nudge_x = 0.3, direction = "y",
                  segment.colour = NA) +
  scale_x_discrete(expand = expansion(add = c(0.5, 1.5))) +
  labs(title = "Quarry blasts by month (grey = years, black = average excl. 2019, red = 2019)",
       x = "Month", y = "Number of blasts")
```
Quarry blasting shows a clear seasonal cycle - higher through summer, lower in winter - most likely driven by summer peaks in construction demand. Against the baseline, **summer 2019** stands out: blasts fell to roughly winter levels when they should have peaked, coinciding with the Ridgecrest sequence in July 2019. Two hard-to-separate mechanisms likely explain this: quarry operators near the sequence may have **paused blasting** in response to earthquakes, or the flood of aftershocks may have **overwhelmed catalogue processing** so blasts were under-detected during that window. 

```{r et-hour}
# Heatmap of event distribution across a week
eq_qb%>%
  count(et, wday, hour) %>%
  complete(et, wday, hour = 0:23, fill = list(n = 0)) %>%
  group_by(et) %>% mutate(frac = n / sum(n)) %>% ungroup() %>%
  ggplot(aes(hour, wday, fill = frac)) +
  geom_tile() +
  facet_wrap(~ et, labeller = as_labeller(et_labs)) +
  scale_y_discrete(limits = rev) +                        
  scale_fill_gradientn(colours = c("black", "darkred", "red", "orange")) +
  labs(title = "When events happen: day-of-week x hour-of-day (local time)",
       x = "Hour (local)", y = "Day of week", fill = "Fraction")
```

The two populations occupy the working week very differently. Earthquakes fill the entire grid roughly evenly, as expected for a natural process without a reason to prefer any hour or day. On the other hand, quarry blasts are sharply concentrated in **weekday daytime working hours** (09:00-17:00 local Monday-Friday), going dark overnight and across the weekend. This is a strong signature of human-scheduling, and is likely a strong blast discriminant. 

### Magnitude

```{r et-mag}
# Kernel density estimates of magnitude distributions
eq_qb%>%
  ggplot(aes(mag, fill = et,colour=et)) +
  geom_density(alpha = 0.3, linewidth = 0.8)+
  scale_colour_manual(name = "Type", values = et_cols,
                    labels = et_labs)+
  scale_fill_manual(name = "Type", values = et_cols,
                    labels = et_labs) +
  labs(title = "Magnitude distribution: earthquakes vs quarry blasts",
       x = "Magnitude", y = "Density", fill = "Type")
```

The kernel density estimates of magnitude distributions are also distinct. Quarry blasts are *tightly packed* at small magnitudes (almost entirely between $0.5$ and $2.5$), reflecting the limited and repeatable energy of a controlled explosion, whereas earthquakes span a *much wider* range (from $-0.5$ to $4$ and beyond), with a long right tail characteristic of the **Gutenberg-Richter** law [@gutenberg1949, p. 17]. In both cases, the fall-off in density below the peak is the signature of catalogue incompleteness: below the magnitude of completeness $M_c$ (which we determine later) progressively more small events escape detection.

### Depth

```{r et-depth}
# Kernel density estimates of depth distributions
eq_qb%>%
  ggplot(aes(depth, fill = et,colour=et)) +
  geom_density(alpha = 0.5) +
  scale_colour_manual(name="Type",values = et_cols,
                    labels = et_labs)+
    scale_fill_manual(name = "Type", values = et_cols,
                    labels = et_labs) +
  labs(title = "Depth distribution: earthquakes vs quarry blasts",
       x = "Depth (km)", y = "Density", fill = "Type")
```
Quarry blasts typically detonate at or just below the surface, so their depths pile up near zero, while earthquakes are distributed over a range of crustal depths.

### Spatial

```{r et-map}
# Examining quarry blast and earthquake locations
eq_qb%>%
    ggplot() + base_map + 
  geom_point(aes(lon, lat, colour = et), size = 0.3, alpha = 0.4) +
  scale_colour_manual(values = et_cols,
                    labels = et_labs) +
  geom_point(data = events, aes(lon, lat, shape = type), size = 3, stroke = 1.2) +
  scale_shape_manual(values = c(Mainshock = 8, Foreshock = 5), name = "Key")+
  coord_quickmap( xlim = c(spatial_window$lon_min - 0.5, spatial_window$lon_max + 0.5),
            ylim = c(spatial_window$lat_min - 0.5, spatial_window$lat_max + 0.5)) +  
  labs(title = "Earthquakes vs quarry blasts (regional context)",
       x = "Longitude", y = "Latitude", colour = "Type") +
  guides(colour = guide_legend(override.aes = list(size = 2, alpha = 1)))
```

The two are also spatially distinct: earthquakes typically trace linear fault structure, while quarry blasts occur in a few tight, isolated clusters at fixed sites - presumably the quarrying sites themselves.

### Decision

The above categories separate quarry blasts from earthquakes. We therefore **remove** all `qb` events and **retain** the `eq` events for modelling. 

```{r make-quakes}
# Retain only earthquake events
cat <- cat_clean %>% filter(et == "eq")
cat("Earthquakes:", nrow(cat), "of", nrow(cat_raw), "total events\n")
```

# Catalogue quality assessment

We explore feature and artefacts of the retained events (earthquakes within the SCSN network).

## Missing values
 
A quick check of how many entries are missing in each field: none.
 
```{r missing}
colSums(is.na(cat))
```

## Event timing artefacts
 
Although we have now filtered our catalogue to earthquake events only, it still remains a possibility that a few **quarry blasts** (controlled explosions) can be mislabelled as earthquakes. Unlike natural earthquakes, blasts happen during working hours on weekdays, so a spike in a particular time-of-day is a red flag for contamination. 
The hour-of-day frequency plot actually reveals a slight drop during the working day. There are a couple of possible explanations for this. The first is **detection asymmetry** - lower night-time cultural noise lowers the detection threshold, revealing more small events [@rydelek1989testing]. The second is a potential artefact - real earthquakes have been mislabelled as blasts and been stripped from the daytime population. An important note here is that events that occurred during the 2019 Ridgecrest sequence below $M2.5$ remain largely unreviewed [@scedc_changehistory], which might contribute to the latter.
```{r hour-of-day}
# Hour of day frequency
ggplot(cat, aes(hour)) +
  geom_bar() +
  labs(title = "Event count by local hour-of-day",
       subtitle = "Local time",
       x = "Local (PST/PDT)", y = "Count")
```
 
We consider events above a magnitude which is likely above $M_c$ to eliminate the influence of time-of-day variations in $M_c$.
Plotting the events above M2.5, we see relative uniformity throughout, with the exception of Friday evenings into Saturday morning. This is likely due to the aftershocks following the $M7.1$ Ridgecrest mainshock on Friday 5th July 2019 at ~20:20 PDT. We additionally see less notable outliers on Thursdays around $11:00$ and Wednesday around $17:00$, although these can be attributed to random noise. This indicates minimal mislabelling above $M2.5$.

```{r hour-by-wday-big}
# Frequency of events above M2.5 by day-of-week and hour-of-day
cat %>%
  filter(mag >2.5) %>%
  count(wday, hour) %>%
  ggplot(aes(hour, wday, fill = n)) +
  geom_tile() + scale_fill_viridis_c() +
  labs(title = "Large events (M > 2.5): hour x weekday",
       x = "Hour (local)", y = NULL, fill = "Count")
```

## Depth artefacts
 
Recall that depth is unknown and it is perhaps the hardest quantity to estimate, so its histogram often shows **artefacts** - spikes at "default" depths (e.g. exactly 10 km) where the location algorithm could not resolve depth and was left at a default value. A 1 km-bin
histogram looks reasonable, with no clustering at $0$ and no dominant spikes at round depths.
 
```{r depth-hist}
# Histogram of depth (overview)
ggplot(cat, aes(x=depth)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Depth histogram (1 km bins)", x = "Depth (km)", y = "Count")
```
To test for finer rounding artefacts directly, we inspect the distribution of the fractional part of the depth. A uniform distribution indicates no preferential reporting at particular decimal values.
```{r depth-fractional}
# Histogram of fractional component of depth
ggplot(cat, aes(depth %% 1)) +
  geom_histogram(binwidth = 0.01, boundary = 0) +
  labs(title = "Fractional part of reported depth",
       x = "Depth mod 1 (km)", y = "Count")
```
Plotting depth against real time shows a stable depth distribution across the catalogue, indicating consistent processing regimes.
 
```{r depth-time}
# Depth time series
ggplot(cat, aes(datetime, depth)) +
  geom_point(size = 0.1, alpha = 0.15) +
  scale_y_reverse() +
  labs(title = "Depth vs date-time (whole catalogue)",
       x = NULL, y = "Depth (km)")
```

Regardless, as depth is not used in the temporal model, location quality only affects our analysis solely through spatial selection into the study region.
 
## Location Quality
`q` is the catalogue's own **location-quality grade** (A = best-located,D/Z = poorly located) [see @scedc_format]. The vast majority of events ($97.3 \%$) are grade *A* (+- 1 km horizontal distance, +- 2 km depth) and a further $1.13 \%$ grade B (+- 2 km horizontal distance, +- 5 km depth), with only $1.49 \%$ at grades C,D, or Z ($ \geq +- 5$km horizontal, no depth bound). The *0.75* category seems to be a data artefact - it is not listed as a valid data entry in the data set description. 

A map of the `q` categories reveals an expected pattern - location quality lowers towards the edges of the network. However, there are examples of events with poor location quality close to the centre of the network, e.g.,  close to the Ridgecrest fore- and main-shock, potentially pointing to short-term aftershock incompleteness (STAI), which we discuss in detail when discussing magnitude of completeness.
 
```{r quality}
# Summary and map of quality categories 
cat %>% count(q) %>% arrange(desc(n))

cat%>%
filter(q != "0.75")%>%
ggplot()+base_map +
  geom_point(aes(lon, lat, colour = q, alpha=q)) +
  scale_colour_manual(values = c(A = "grey", B = "black",
                        C = "orange", D = "red", Z = "blue")) +
  geom_point(data = events, aes(lon, lat, shape = type), size = 3, stroke = 1.2) +
  scale_shape_manual(values = c(Mainshock = 8, Foreshock = 5), name = "Key")+
  scale_alpha_manual(values = c(A = 0.1, B = 0.4, C = 0.9, D = 0.9, Z = 0.9),
                     guide = "none") +
  scale_size_manual(values = c(A = 0.3, B = 0.4, C = 0.9, D = 1, Z = 1),
                    guide = "none") +
 coord_quickmap(xlim = range(cat$lon), ylim = range(cat$lat))+
  labs(colour = "Location grade",
       title = "Events by location-quality grade", x = "longitude", y = "latitude")
```

We also observe a magnitude dependence. The majority of hard-to-locate events are of a lower magnitude.

```{r}
# Magnitude density by quality
cat %>%
  filter(q != "0.75")%>%
  mutate(q = factor(q, levels = c("A","B","C","D","Z"))) %>%
  ggplot(aes(mag, colour = q)) +
  geom_density() +
  scale_colour_manual(
    values = c(A = "grey", B = "black", C = "orange", D = "red", Z = "blue"),
    name   = "Location grade"
  ) +
  labs(title = "Magnitude density by location-quality grade",
       x = "Magnitude", y = "Density")
```

Switching the relation, we see that as magnitude of the event increases from negative to $0$, the proportion of A events increase, reflecting the fact that more sensors can pick up the earthquake. On the other hand, as we go past $M0$, this figure starts to decrease, perhaps due to a rupturing of extended sources rather than points, leading to a hard-to-identify hypocentre.
```{r}
# Location grade composition by magnitude bin
cat %>%
filter(q != "0.75")%>%
  mutate(q       = factor(q, levels = c("A","B","C","D","Z")),
         mag_bin = floor(mag)) %>%
  count(mag_bin, q) %>%
  ggplot(aes(mag_bin, n, fill = q)) +
  geom_col(position = "fill") +                       
  scale_y_continuous(labels = scales::percent) +       
  scale_fill_manual(
    values = c(A = "grey70", B = "grey40", C = "orange", D = "red", Z = "blue"),
    name   = "Location grade"
  ) +
  labs(title = "Location-grade composition by magnitude bin",
       x = "Magnitude (1-unit bins)", y = "Percentage of events in bin")
```

## Magnitude reporting artefacts

**Natural time** plots each event evenly spaced along the x-axis by its order,
rather than its date. This stretches out busy periods so that structure hidden in
dense clusters becomes visible. Here, there is not much evidence for artefacts 
such as horizontal *banding* from magnitude rounding (at any time),
or a spurious pile-up at magnitude 0. The most interesting point is
**time varying sensitivity**. Following the largest event, we observe a sharp 
increase in the lowest detected magnitude from $\sim 0$ to $\sim 2$, before
gradual recovery, pointing to the aforementioned STAI.
 
```{r mag-natural}
# Magnitude vs event index
ggplot(cat, aes(event_num, mag)) +
  geom_point(size = 0.1) +
  labs(title = "Magnitude vs natural time (event index)",
       x = "Event number", y = "Magnitude")
```
 
The absence of $M0$ clustering and the $0.01$ magnitude rounding are confirmed
by a very fine histogram. The distribution appears near continuous and without a
spike at 0.
 
```{r mag-zero}
# Magnitude histogram (fine bins)
ggplot(cat, aes(mag)) +
  geom_histogram(binwidth = 0.01, boundary = 0) +
  coord_cartesian(xlim = c(-0.5, 3)) +
  labs(title = "Fine magnitude histogram (0.01 bins): rounding artefacts",
       x = "Magnitude", y = "Count")
```

# Magnitude completeness
## Frequency-magnitude and completeness

The magnitude distribution of the catalogue is described by the Gutenberg-Richter law [@gutenberg1944frequency],

$$
\log_{10}N_M=a-bM,
$$
where $N_M$ is the nymber of events of magnitude at least $M$, the $b$-value controls the relative frequency of large to small events, and the $a$-value represents the intercept (number of events of non-negative magnitude). We estimate $b$ by a maximum- likelihood estimator [@aki1965maximum,@utsu1965method], named the Aki-Utsu estimator. Applying the correction for magnitude binning [@bender1983maximum], it reads:
$$
\hat{b} = \frac{\log_{10} e}{\bar{M} - \left(M_c - \dfrac{\Delta M_{\text{bin}}}{2}\right)},
$$
where $\bar{M}$ is the mean magnitude of events above the magnitude threshold $M_c$ and $\Delta M_{bin}$ is the catalogue's magnitude bin width ($0.1$ here).

The a-value follows by substituting $\hat{b}$ into the Gutenberg-Richter law,
$$
\hat{a}=\log_{10}N_{M_c}+\hat{b}M_c.
$$

### Completeness magnitude

Note that the expressions above require $M_c$ - the magnitude of completeness above which all events are included.

There are two primary subcategories of methods to determine magnitude of completeness. The first is network-based methods, which use station properties such as locations, detection probabilities, and network performance to inform magnitude of completeness. Although methods in this category incorporate more information in calculations of magnitude of completeness, they are complex and require large datasets. 

As such, we focus on the other category: deriving $M_c$ directly from the catalogue. Here, we consider the three main methods in this category:

- **Maximum curvature (MAXC)**. Point of maximum curvature on frequency-magnitude distribution (FMD) [@wiemer2000minimum]. A typical FMD of the catalogue will reveal three regions: incompleteness at low magnitudes (small events undetected),
a straight log-linear part (good GR behaviour), and scatter at high magnitudes (too few big events for stable counts). The magnitude where the straight part begins is the **completeness magnitude $M_c$**. @woessner2005assessing suggests finding the magnitude bin with the highest frequency of events in the non-cumulative frequency-magnitude distribution.

- **Goodness-of-fit test (GFT)**. The observed FMD is compared against synthetic Gutenberg-Richter distributions generated from the maximum-likelihood a- and b-values at each trial cutoff ($M_{co}$), with $M_c$ taken as the lowest magnitude at which the power law reproduces a fixed percentage: 

$$
R(a, b, M_{co}) = 100 \left(1 - \frac{\sum_{i:M_i \geq M_{co}} |B_i - S_i|}{\sum_i B_i}\right),
$$

where $B_i$ and $S_i$ are the observed and predicted cumulative
number of events in each magnitude bin. Here we consider $90\%$, as $95 \%$ is rarely obtained for real catalogues [@wiemer2000minimum].

- **b-value stability method (MBS)**. Based on the assumption that $b$-values ascend for $M_{co} < M_c$, remain constant for $M_{co} \geq M_c$, and ascend again for $M_{co} >> M_c$. With $\delta b=2.3 b^2 \sqrt{\frac{\sum_{i=1}^N(M_i-\bar{M})^2}{N(N-1)}}$ (the Shi-Bolt [@shibolt1982standard] uncertainty of the Aki b-value MLE [@aki1965maximum] estimate), $M_c$ is defined as the lowest $M_{co}$ where $|b_{ave}-b| < \delta b$, where $b_{ave}=\sum_{M_{i=co}}^{M_{co}+5dm} \frac{\hat{b}(M_{i})}{5}$[@woessner2005assesing], where $dm$ is the bin-size.

### Whole catalogue $M_c$

We utilise each of the aforementioned methods to provide a conservative estimate for the whole-catalogue magnitude of completeness, to provide better-informed selection of the Ridgecrest sequence.

Both MAXC and GFT estimate a catalogue-wide completeness magnitude of $M_c = 0.8$, whereas MBS provides a more conservative $M_c = 1.8$. Since the present objective is to define the Ridgecrest sequence rather than estimate the GR parameters, we adopt $M_c = 0.8$. 
```{r catalog-mc}
# Whole catalogue magnitude of completeness 
# Frequency-magnitude data
gr <- gr_data(cat$mag, dm)

# MAXC and GFT estimates
mc_maxc <- estimate_mc_maxc(cat$mag, dm)
mc_gft_val <- mc_gft(cat$mag, dm)$mc

# Plot FMD and overlay MAXC and GFT estimates
ggplot(gr, aes(mag)) +
  geom_col(aes(y = incremental),fill = "grey75", colour = "black", width = dm) +
  geom_line(aes(y = cumulative),colour = "steelblue",linewidth = 1) +
  scale_y_log10() +
  geom_vline(xintercept = mc_maxc, colour = "red", linetype = "dashed", 
             linewidth = 0.8) +
  geom_vline(xintercept = mc_gft_val, colour = "darkblue", linetype = "dashed",
             linewidth = 0.8) +
  annotate("text",  x = mc_maxc, y = max(gr$cumulative), 
           label = paste0("MAXC = ", mc_maxc), colour = "red",
           hjust = -0.05, vjust = -0.3) +
  annotate("text", x = mc_gft_val,  y = max(gr$cumulative)/3, 
           label = paste0("GFT = ", mc_gft_val),
           colour = "darkblue", hjust = 1.5) +
  labs(title = "Frequency-magnitude distribution", x = "Magnitude",
    y = "Number of events (log scale)"
  )

# Plot b-values for each Mc
mbs <- mc_mbs(cat$mag, dm)
ggplot(mbs$curve, aes(mc, b)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.2) +
  geom_errorbar(
    aes(ymin = b - b_err,
        ymax = b + b_err),
    width = 0.03
  ) +
  geom_vline(xintercept = mbs$mc, colour = "forestgreen", linetype = "solid") +
  geom_vline(xintercept = mc_maxc, colour = "red", linetype = "dashed") +
  geom_vline(xintercept = mc_gft_val, colour = "darkblue", linetype = "dashed") +
  annotate("text",  x = mc_maxc, y = max(mbs$b), 
           label = paste0("MAXC = ", mc_maxc), colour = "red",
           hjust = -0.05, vjust = -0.3) +
  annotate("text", x = mc_gft_val,  y = max(mbs$b)/3, 
           label = paste0("GFT = ", mc_gft_val),
           colour = "darkblue", hjust = 1.5, vjust = -1) +
  annotate("text",  x = mbs$mc, y = max(mbs$b), 
           label = paste0("MBS = ", mbs$mc), colour = "forestgreen",
           hjust = -0.5, vjust = -0.3) +
  labs(title = "b-value stability", x = expression(M[c]), y = expression(hat(b))
  )
```

### Completeness drift

We use GFT and MAXC to estimate magnitude of completeness ($M_c$) *within each year*. This checks for **completeness drift**, where the network gets better at *detecting* small earthquakes over time.

The plot indicates that $M_c$ is essentially *stationary* across the decade, meaning there is no progressive detection improvement to correct for. While it may be tempting to consider a $M_c$ of $1-1.2$ going forwards, these are annual means and so blur any sub-annual structure, in particular the STAI that arises when a large mainshock and its early aftershocks obscure smaller events and can saturate nearby stations. This effect is already visible even at this coarse resolution: for both methods, $M_c$ sharply rises between $2018$ and $2019$ (when the Ridgecrest mainshock occured).
```{r annual-mc}
# Magnitude of completeness over time (determined by MAXC and GFT)
cat %>%
  group_by(year) %>%
  summarise(n= n(),
            mc_maxc = estimate_mc_maxc(mag, dm),
            mc_gft  = mc_gft(mag, dm)$mc,
            .groups = "drop") %>%
  pivot_longer(c(mc_maxc, mc_gft), names_to = "method",
                      values_to = "mc") %>%
  ggplot(aes(year, mc, colour = method)) +
  geom_line() + geom_point(size = 1) +
  scale_x_continuous(breaks = seq(2016, 2025, 3)) +
  scale_colour_discrete(labels = c(mc_maxc = "MAXC", mc_gft = "GFT"))+
  labs(title = "Completeness magnitude by year (MAXC vs GFT)",
       subtitle = "No clear trend in completeness",
       x = "Year", y = "Mc", colour = "Method")
```

### Criticisms

There are naturally criticisms of this approach, such as the right-tail of the distribution, which implies the possibility of events such as $M10$ earthquakes, which to this day have not yet occurred. As such, there have been many advancements made to the GR law since its first implementation in $1949$, such as the **tapered** GR law [@kagan2002seismic], yet it remains the standard for earthquake modelling, particularly where we avoid overcomplicating models.

More recent estimators are designed to be robust to STAI that raises $M_c$ after large events - most notably the $b$-positive estimator of @vanderelst2021bpositive, which estimates $b$ from the positive differences between successive magnitudes. We nonetheless stick with the Aki-Utsu estimator for simplicity.

# Sequence selection

Having observed whole catalogue properties, we move onto consideration of the Ridgecrest sequence. To this end, we set the spatial and temporal windows that define the sequence, based on events in the catalogue with magnitude above the loose catalogue $M_c = 0.8$. 
## Spatial window 
 
We consider a hex plot of all events post the mainshock, which provides density of event occurrence. This reveals areas of high seismicity - in particular to the northwest and southwest of the Ridgecrest mainshock. Additionally, we observe high seismicity stretching from the centre of the catalogue towards the Mexican border. The purple/black region in between this region and the Ridgecrest region highlights unrelated triggering - California displays complex fault geometry, so these events are likely a result of a different earthquake.
 
```{r cat-heatmap}
# Retain only events with magnitude above 0.8 in determination of the spatiotemporal window.

cat_comp <- cat %>% filter(mag >= 0.8)
cat_comp %>%
  filter(datetime > mainshock_time) %>%
  ggplot(aes(lon, lat)) +
  base_map +
  geom_hex(bins = 1000) +                                  
  geom_point(data = events, aes(lon, lat, shape = type),
             size = 3, colour = "cyan", stroke = 1.2) +   
  scale_shape_manual(values = c(Mainshock = 8, Foreshock = 5), name = "Key") +
scale_fill_viridis_c(option = "inferno", trans = "log10", name = "Count") +  
  coord_quickmap(xlim = range(cat$lon), ylim = range(cat$lat)) +
  labs(title = "Aftershock density (log colour scale)",
       x = "Longitude", y = "Latitude")
```

The `ETAS.inlabru` method supports rectangular regions only. Zooming in and increasing the resolution of the hex map, we overlay three candidate spatial windows. The red window, used by @bayliss2020data has corners $[35^\circ N, -118^\circ E], [37^\circ N, -117^\circ E]$. This paper was interested in characterising the background seismicity on a declustered Ridgecrest catalogue, and was therefore able to consider a more conservative region. The dark blue window, used by @vanderelst2021bpositive,  has corners $[35.2^\circ N, -118.2^\circ E], [36.4^\circ N, -117^\circ E]$. We extend this box further north (up to $36.6^\circ N$) for the green window, which includes further events that seem Ridgecrest-related.

```{r spatial-window}
# Candidate spatial windows for Ridgecrest aftershock zone
# Data frame for windows
box_data <- data.frame(
  xmin = c(-118, -118.2, -118.2),
  xmax = c(-117, -117, -117),
  ymin = c(35, 35.2, 35.2),
  ymax = c(37, 36.4, 36.6),
  window = c("Bayliss et al. (2020)", "van der Elst (2021)", "Proposed window") 
)

box_data$window <- factor(box_data$window, levels = c(
  "Bayliss et al. (2020)", 
  "van der Elst (2021)", 
  "Proposed window"
))

# Plot
cat_comp %>%
  filter(datetime > mainshock_time) %>%
  ggplot(aes(lon, lat)) +
  base_map +
  geom_hex(bins = 8000) +                             
  geom_point(data = events, aes(lon, lat, shape = type),
             size = 3, colour = "cyan", stroke = 1.2) +
  geom_rect(data = box_data, 
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, colour = window),
            inherit.aes = FALSE, fill = NA, linewidth = 0.5) +
  scale_colour_manual(
    name = "Windows",
    values = c("Bayliss et al. (2020)" = "red", "van der Elst (2021)" = "darkblue", "Proposed window" = "forestgreen")
  ) +
  scale_shape_manual(values = c(Mainshock = 8, Foreshock = 5), name = "Key") +
  scale_fill_viridis_c(option = "inferno", trans = "log10", name = "Count") +
  coord_quickmap(xlim = c(mainshock_lon - 1.5, mainshock_lon + 1.5), 
                 ylim = c(mainshock_lat - 1.5, mainshock_lat + 1.5)) +
  labs(title = "Aftershock density (log colour scale)",
       x = "Longitude", y = "Latitude")
```

```{r filter-spatial}
# Restrict spatial window, assists in determining temporal window
cat_spatial <- cat_comp%>%filter(lon >= -118.2, lon <= -117, lat >= 35.2, lat <= 36.6)
```

## Temporal Window

We plot cumulative events over time to provide an initial assessment of the temporal evolution of seismicity, including:
- When the quiescent period begins prior to the foreshock
- The duration over which elevated seismicity can reasonably be attributed to the mainshock.

The former is relatively trivial, whereas defining the latter involves more work. @hainzl2016 defines the apparent aftershock duration as the time required for the aftershock rate to decay to the background seismicity rate. Here, background seismicity is estimated from the average seismicity rate observed prior to the $M6.4$ foreshock (@shiratori1925sagami, as reviewed by @utsu1995, yielding a rate of $2.56$ events/day.

We first plot a cumulative event count to illustrate the evolution of the sequence (note this approach does not account dor temporal variations in catalogue completeness and considers the previously derived whole-catalogue $M_c$). To assess the transition towards background-like activity, daily event rate are calculated and smoothed using a trailing 90-day rolling mean evaluated at quarterly intervals. Rather than defining sequence termination solely as the first point at which the rate reaches the background level, we identify the point at which seismicity has approached and stabilised near the background rate while avoiding potential contamination from later independent seismic sequences.

Based on this assessment, 1 July 2023 is selected as the end of the Ridgecrest aftershock analysis window. By this time, the preceding 90-day mean seismicity rate has largely stabilised, although it remains elevated relative to the estimated pre-foreshock background rate of 2.56 events/day. This indicates that the rapid decay phase of the triggered sequence has largely concluded, while retaining the later-time decay tail required for analysis of the temporal evolution of aftershock productivity. The selected window provides approximately four years of post-mainshock observations and 3.5 years of pre-foreshock activity for comparison and estimation of background seismicity. Although longer background catalogues could further constrain the regional seismicity rate, the available interval provides a consistent reference period for this analysis [@naylor2023].

```{r temporal-window}
### Identifying temporal window - evaluating rolling mean event counts quarterly
# Add cumulative count and days since foreshock columns (negative for events before foreshock)
cat_spatial_cumulative <- cat_spatial %>%
  arrange(datetime) %>%
  mutate(cumulative_count = row_number(),
    days_from_foreshock = days_since(datetime, foreshock_time)
)

### Calculate background seismicity rate

# Retain only events prior to the foreshock (activity heightened after foreshock)
pre_foreshock <- cat_spatial_cumulative %>%
  filter(days_from_foreshock < 0)

# Slope = Total events in window / Total days in window
baseline_slope <- nrow(pre_foreshock) /  max(abs(pre_foreshock$days_from_foreshock))

print(paste("Baseline background rate:", round(baseline_slope, 3), "events/day"))

### Post-mainshock daily rate

# 3-month checkpoints after mainshock
checkpoints <- tibble(
  date = seq(
    from = as.Date("2019-10-01"),
    to   = post_end,
    by   = "3 months"
  )
)

# Average 30-day rolling rates around each checkpoint
checkpoint_rates <- checkpoints %>%
  rowwise() %>%
  mutate(
    mean_rolling_rate = mean(
      post_daily$rolling_rate[
        post_daily$date > date %m-% months(3) &
        post_daily$date <= date
      ],
      na.rm = TRUE
    )
  ) %>%
  ungroup()

ggplot(checkpoint_rates, aes(date, log(mean_rolling_rate))) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_hline(
    yintercept = log(baseline_slope),
    linetype = "dashed",
    colour = "red"
  ) +
  labs(
    title = "Post-mainshock seismicity decay",
    x = "Checkpoint date",
    y = "Mean 30-day rolling event rate \n
    (preceding 3 months), log-scale"
  )
```
```{r cumulative-visualisation}
# Visualisation of cumulative seismicity plots
sequence_end <- as.POSIXct("2023-07-01", tz = "UTC")

ggplot(cat_spatial_cumulative,
       aes(datetime, cumulative_count)) +
  geom_step(colour = "black", linewidth = 0.8) +
  geom_vline(xintercept = foreshock_time,  colour = "darkblue",
             linetype = "dashed") +
  geom_vline(xintercept = mainshock_time, colour = "red",
             linetype = "dashed") +
  geom_vline(xintercept = sequence_end,  colour = "forestgreen",
             linetype = "dashed") +
  annotate( "text", x = foreshock_time, y = max(cat_spatial_cumulative$cumulative_count), label = "Foreshock",
    colour = "darkblue",hjust = -0.1) +
  annotate("text", x = mainshock_time,y = 0.95 * max(cat_spatial_cumulative$cumulative_count), label = "Mainshock",
    colour = "red",hjust = -0.1) +
  annotate("text", x = sequence_end, y = 0.90 * max(cat_spatial_cumulative$cumulative_count), label = paste( "Sequence end\n",  format(sequence_end, "%Y-%m-%d") ), colour = "forestgreen",hjust = -0.1,
           vjust = 1 ) +
  labs(
    title = "Cumulative seismicity through time",
    x = "Date",
    y = "Cumulative number of earthquakes"
  )
```

```{r filter-temporal}
# Restrict to chosen spatio-temporal region. Bring back events below whole
# catalogue Mc for rigorous sequence Mc and b-value determination
cat_ridgecrest <- extract_sequence(cat,lon_range = c(-118.2, -117),lat_range = c(35.2, 36.6),t_range = c(min(cat$datetime), sequence_end))
```

# $M_0$ and $b$


Having determined the spatio-temporal window for the Ridgecrest sequence, we know determine the fixed modelling parameters - namely $M_0$ and the Gutenberg-Richter $b$-value. Recall that for the entire catalogue both MAXC and GFT estimate a catalogue-wide completeness magnitude of $M_c = 0.8$, whereas MBS provides a more conservative $M_c = 1.8$. 

```{r sequencce-mc}
# Whole sequence magnitude of completeness 
# Frequency-magnitude data
gr_ridgecrest <- gr_data(cat_ridgecrest$mag, dm)

# MAXC and GFT estimates
mc_maxc <- estimate_mc_maxc(cat_ridgecrest$mag, dm)
mc_gft_val <- mc_gft(cat_ridgecrest$mag, dm)$mc

# Plot FMD and overlay MAXC and GFT estimates
ggplot(gr_ridgecrest, aes(mag)) +
  geom_col(aes(y = incremental),fill = "grey75", colour = "black", width = dm) +
  geom_line(aes(y = cumulative),colour = "steelblue",linewidth = 1) +
  scale_y_log10() +
  geom_vline(xintercept = mc_maxc, colour = "red", linetype = "dashed", 
             linewidth = 0.8) +
  geom_vline(xintercept = mc_gft_val, colour = "darkblue", linetype = "dashed",
             linewidth = 0.8) +
  annotate("text",  x = mc_maxc, y = max(gr_ridgecrest$cumulative), 
           label = paste0("MAXC = ", mc_maxc), colour = "red",
           hjust = -0.05, vjust = -0.3) +
  annotate("text", x = mc_gft_val,  y = max(gr_ridgecrest$cumulative)/3, 
           label = paste0("GFT = ", mc_gft_val),
           colour = "darkblue", hjust = 1.5) +
  labs(title = "Frequency-magnitude distribution", x = "Magnitude",
    y = "Number of events (log scale)"
  )

# Plot b-values for each Mc
mbs <- mc_mbs(cat_ridgecrest$mag, dm)
ggplot(mbs$curve, aes(mc, b)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.2) +
  geom_errorbar(
    aes(ymin = b - b_err,
        ymax = b + b_err),
    width = 0.03
  ) +
  geom_vline(xintercept = mbs$mc, colour = "forestgreen", linetype = "solid") +
  geom_vline(xintercept = mc_maxc, colour = "red", linetype = "dashed") +
  geom_vline(xintercept = mc_gft_val, colour = "darkblue", linetype = "dashed") +
  annotate("text",  x = mc_maxc, y = max(mbs$b), 
           label = paste0("MAXC = ", mc_maxc), colour = "red",
           hjust = -0.05, vjust = -0.3) +
  annotate("text", x = mc_gft_val,  y = max(mbs$b)/3, 
           label = paste0("GFT = ", mc_gft_val),
           colour = "darkblue", hjust = 1.5, vjust = -1) +
  annotate("text",  x = mbs$mc, y = max(mbs$b), 
           label = paste0("MBS = ", mbs$mc), colour = "forestgreen",
           hjust = -0.5, vjust = -0.3) +
  labs(title = "b-value stability", x = expression(M[c]), y = expression(hat(b))
  )
```

While it's tempting to determine the $M_0$ from the entire catalogue $M_c$, we must consider short-term aftershock incompleteness (STAI). This effect is observed in the below graph of weekly $M_c$ - note the increase across all methods for $M_c$ values post fore- and mainshock: there is no single value that describes the catalogue.
```{r rolling-mc}
# Rolling Mc - completeness threshold increases in wake of foreshock/mainshock activity
cat_ridgecrest %>%
  filter(datetime >= as.POSIXct("2019-05-30", tz = "UTC"),
         datetime <= as.POSIXct("2023-07-01", tz = "UTC")) %>%
  arrange(datetime) %>%
  mutate(window_date = cut(datetime,
                      breaks = seq(as.POSIXct("2019-05-30", tz = "UTC"),
                                   as.POSIXct("2023-07-02", tz = "UTC"),
                                   by = "7 days"),labels = FALSE) %>% 
    as.numeric(),
    window_label = as.POSIXct("2019-05-30", tz = "UTC") + (window_date - 1 + 0.5) * 7 * 86400  
  ) %>%
  group_by(window_label) %>%
 summarise(n = n(),mc_gft = mc_gft(mag, dm, min_n = 20)$mc,
          mc_maxc = estimate_mc_maxc(mag, dm), mc_mbs = mc_mbs(mag, dm)$mc,.groups = "drop") %>%
  filter(n >= 20) %>%
  pivot_longer(c(mc_gft, mc_maxc, mc_mbs), names_to = "method", values_to = "mc") %>%
  ggplot(aes(window_label, mc, colour = method, linetype = method)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2, alpha = 0.7) +
  geom_vline(xintercept = foreshock_time, linetype = "dashed", 
             colour = "darkblue", linewidth = 0.8, alpha = 0.5) +
  geom_vline(xintercept = mainshock_time, linetype = "dashed", 
             colour = "red", linewidth = 0.8, alpha = 0.5) +
  annotate("text", x = foreshock_time, y = Inf, label = "M6.4", 
           colour = "darkblue", hjust = -0.2, vjust = 1.2, size = 3.5) +
  annotate("text", x = mainshock_time, y = Inf, label = "M7.1", 
           colour = "red", hjust = -0.2, vjust = 1.2, size = 3.5) +
  scale_colour_manual(values = c(mc_gft = "steelblue", mc_maxc = "orange", 
                                 mc_mbs = "forestgreen"),
                      labels = c(mc_gft = "GFT", mc_maxc = "MAXC", 
                                 mc_mbs = "MBS")) +
  scale_linetype_manual(values = c(mc_gft = "solid", mc_maxc = "dashed",
                                   mc_mbs = "dashed"),
                        labels = c(mc_gft = "GFT", mc_maxc = "MAXC",
                                   mc_mbs = "MBS")) +
  labs(title = "Magnitude of completeness: End of May – End of June 2023 (7-day windows)",
       x = "Date", y = "Mc",
       subtitle = "Weekly windows show Mc spike post-mainshock and recovery",
       colour = "Method", linetype = "Method")
```
Zooming into a daily resolution reveals further details for the week beginning on the date of the foreshock - the MAXC method suggests a peak magnitude of completeness of $2$, while MBS provides a much harsher $3.3$, both on the day of the mainshock (6th July 2019). GFT returns NA, indicating a significant mismatch between empirical and theoretical (Gutenberg-Richter) magnitudes.

```{r mainshock-mc}
# Daily magnitude of completeness
cat_ridgecrest %>%
  filter(datetime >= as.POSIXct("2019-07-04", tz = "UTC"),
         datetime <= as.POSIXct("2019-07-10", tz = "UTC")) %>%
  arrange(datetime) %>%
  mutate(window_date = cut(datetime,
                      breaks = seq(as.POSIXct("2019-07-04", tz = "UTC"),
                                   as.POSIXct("2019-07-11", tz = "UTC"),
                                   by = "1 day"),labels = FALSE) %>% 
    as.numeric(),
    window_label = as.POSIXct("2019-07-04", tz = "UTC") + (window_date - 1) * 86400 
  ) %>%
  group_by(window_label) %>%
 summarise(n = n(),mc_gft = mc_gft(mag, dm, min_n = 20)$mc,
          mc_maxc = estimate_mc_maxc(mag, dm), mc_mbs = mc_mbs(mag, dm)$mc,.groups = "drop") %>%
  filter(n >= 20) %>%
  pivot_longer(c(mc_gft, mc_maxc, mc_mbs), names_to = "method", values_to = "mc") %>%
  ggplot(aes(window_label, mc, colour = method, linetype = method)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2, alpha = 0.7) +
  geom_vline(xintercept = foreshock_time, linetype = "dashed", 
             colour = "darkblue", linewidth = 0.8, alpha = 0.5) +
  geom_vline(xintercept = mainshock_time, linetype = "dashed", 
             colour = "red", linewidth = 0.8, alpha = 0.5) +
  annotate("text", x = foreshock_time, y = Inf, label = "M6.4", 
           colour = "darkblue", hjust = -0.2, vjust = 1.2, size = 3.5) +
  annotate("text", x = mainshock_time, y = Inf, label = "M7.1", 
           colour = "red", hjust = -0.2, vjust = 1.2, size = 3.5) +
  scale_colour_manual(values = c(mc_gft = "steelblue", mc_maxc = "orange", 
                                 mc_mbs = "forestgreen"),
                      labels = c(mc_gft = "GFT", mc_maxc = "MAXC", 
                                 mc_mbs = "MBS")) +
  scale_linetype_manual(values = c(mc_gft = "solid", mc_maxc = "dashed",
                                   mc_mbs = "dashed"),
                        labels = c(mc_gft = "GFT", mc_maxc = "MAXC",
                                   mc_mbs = "MBS")) +
  labs(title = "Magnitude of completeness: July 4th - 10th 2019 (1-day window)",
       x = "Date", y = "Mc",
       subtitle = "Daily windows show Mc spike",
       colour = "Method", linetype = "Method")
```

We explore this departure from the GR further for the 6th July (the date of the mainshock). The peak frequency bin is at $M2$, however, we see a relatively flat GR curve between approximately $1.5-2.5$ - this departs from standard empirical observations. The lack of clear exponential drop-off at any magnitude explains why the GFT could not be found - for no cutoff magnitude did empirical and theretical Gutenberg-Richter distributions look similar. This effect is largely resolved by one week after the mainshock (13th July).

```{r mainshock-gr}
# FMD for selected dates - recovery
july6 <- cat_ridgecrest %>%
  filter(datetime >= as.POSIXct("2019-07-06", tz = "UTC"),
         datetime < as.POSIXct("2019-07-07", tz = "UTC"))

july8 <- cat_ridgecrest %>%
  filter(datetime >= as.POSIXct("2019-07-08", tz = "UTC"),
         datetime < as.POSIXct("2019-07-09", tz = "UTC"))

july13 <- cat_ridgecrest %>%
  filter(datetime >= as.POSIXct("2019-07-13", tz = "UTC"),
         datetime < as.POSIXct("2019-07-14", tz = "UTC"))

gr_data(july6$mag)%>%
  filter(cumulative > 0, incremental > 0)%>%
  ggplot(aes(mag)) +
  geom_col(aes(y = incremental),fill = "grey75", colour = "black", width = dm) +
  geom_line(aes(y = cumulative),colour = "steelblue",linewidth = 1) +
  scale_y_log10()+
  labs(title = "Frequency-magnitude distribution (6/7/2019)", x = "Magnitude",
    y = "Number of events (log scale)"
  )

gr_data(july8$mag)%>%
  filter(cumulative > 0, incremental > 0)%>%
  ggplot(aes(mag)) +
  geom_col(aes(y = incremental),fill = "grey75", colour = "black", width = dm) +
  geom_line(aes(y = cumulative),colour = "steelblue",linewidth = 1) +
  scale_y_log10()+
  labs(title = "Frequency-magnitude distribution (8/7/2019)", x = "Magnitude",
    y = "Number of events (log scale)"
  )

gr_data(july13$mag)%>%
  filter(cumulative > 0, incremental > 0)%>%
  ggplot(aes(mag)) +
  geom_col(aes(y = incremental),fill = "grey75", colour = "black", width = dm) +
  geom_line(aes(y = cumulative),colour = "steelblue",linewidth = 1) +
  scale_y_log10()+
  labs(title = "Frequency-magnitude distribution (12/7/2019)", x = "Magnitude",
    y = "Number of events (log scale)"
  )
```

This motivates exploration of time-varying magnitude of completeness. This has been implemented in frequentist frameworks \citep{Helmstetter2006} and more recently in Bayesian ETAS models via rate-dependent incompleteness corrections \citep{Kamranzad2025}. However, the latter approach lacks publicly-available documentation for the inlabru implementation and introduces methodological complexity beyond the scope of this work. We therefore adopt a fixed $M_c$ threshold, acknowledging this as consideration for future refinement.

The above methods validate $M_0 = 2.5$ - while $M_0 = 2.0$ seems attractive (being the peak daily value MAXC estimate of completeness), we recall that MAXC is a conservative estimate. This is further validated through peak daily MBS estimate of $M_c = 3.3$. We pick $2.5$ as the middle ground between both estimates - going any higher risks a significant reduction in available data to fit the model, while going any lower risks introduction of bias into posteriors (in particular misfitting of decay laws -- overestimation of $p$ for Omori--Utsu, underestimation of aftershock productivity of strong earthquakes, overestimation of $c$) due to STAI @hainzl2022, @kagan2004short. 

```{r mc-counts}
# Number of available events at each magnitude cutoff
for (m in c(1, 1.5, 2, 2.5, 3, 3.3)) {
  cat("M0 =", m, ":", sum(cat_ridgecrest$mag >= m, na.rm = TRUE), "events\n")
}
```

The corresponding $b$-value from the Aki-Utsu MLE is $0.79 \pm 0.016$ on 3,336 events.

```{r b-value}
# Filter to observations above cutoff and compute b-value MLE for whole sequence
cat_fit <- cat_ridgecrest %>% filter (mag >= 2.5, !is.na(mag))
b_aki(cat_fit$mag, mc = 2.5, dm = 0.1)
```

```{r b-value-pre}
# Filter to observations above cutoff and compute b-value MLE for pre-foreshock
cat_fit_background <- cat_ridgecrest %>% filter (mag >= 2.5, 
                                        datetime < foreshock_time, !is.na(mag))
b_aki(cat_fit_background$mag, mc = 2.5, dm = 0.1)
```

```{r b-value-post}
# Filter to observations above cutoff and compute b-value MLE for post-mainshock
cat_fit_active <- cat_ridgecrest %>% filter (mag >= 2.5, 
                                        datetime >= mainshock_time, !is.na(mag))
b_aki(cat_fit_active$mag, mc = 2.5, dm = 0.1)
```

# Visualising the sequence

```{r}
g_ou <- function(t, c1, p1) (1 + t / c1)^(-p1)
 
g_ms <- function(t, c2, lam2, b2) {
  (1 + t / c2)^(b2 - 1) * exp(-lam2 * ((c2 + t)^b2 - c2^b2))
}
 
g_rs <- function(t, B, ta) {
  e <- exp(-t / ta)
  (1 - B) * e / (1 - B * e)
}
 
## Total mass  G_k(0, Inf)
mass_ou <- function(c1, p1)        c1 / (p1 - 1)
mass_ms <- function(c2, lam2, b2)  c2^(1 - b2) / (lam2 * b2)
mass_rs <- function(B, ta)        -(1 - B) * ta * log(1 - B) / B
 
## Normalised densities 
f_ou <- function(t, c1, p1)       g_ou(t, c1, p1)       / mass_ou(c1, p1)
f_ms <- function(t, c2, lam2, b2) g_ms(t, c2, lam2, b2) / mass_ms(c2, lam2, b2)
f_rs <- function(t, B, ta)        g_rs(t, B, ta)        / mass_rs(B, ta)
 
## Distribution functions
## log1p(-x) = log(1 - x), used for numerical stability as B -> 1.
F_ou <- function(t, c1, p1) 1 - (1 + t / c1)^(1 - p1)
 
F_ms <- function(t, c2, lam2, b2) -expm1(-lam2 * ((c2 + t)^b2 - c2^b2))
 
F_rs <- function(t, B, ta) 1 - log1p(-B * exp(-t / ta)) / log1p(-B)
 
## Parameters  
 
par_ou <- list(c1 = 0.011,  p1   = 1.12)
par_ms <- list(c2 = 0.0004, lam2 = 1.01, b2 = 0.22)
par_rs <- list(B  = 0.99998, ta  = 188)
 
kernel_levels <- c("Omori--Utsu", "Modified stretched exponential", "Rate-state")

## Evaluate on a logarithmic time grid
t_grid <- 10^seq(-4, 4, length.out = 2000)   # days
 
kern <- bind_rows(
  tibble(t = t_grid,
         kernel   = kernel_levels[1],
         density  = f_ou(t_grid, par_ou$c1, par_ou$p1),
         cumulative = F_ou(t_grid, par_ou$c1, par_ou$p1)),
  tibble(t = t_grid,
         kernel   = kernel_levels[2],
         density  = f_ms(t_grid, par_ms$c2, par_ms$lam2, par_ms$b2),
         cumulative = F_ms(t_grid, par_ms$c2, par_ms$lam2, par_ms$b2)),
  tibble(t = t_grid,
         kernel   = kernel_levels[3],
         density  = f_rs(t_grid, par_rs$B, par_rs$ta),
         cumulative = F_rs(t_grid, par_rs$B, par_rs$ta))
) |>
  mutate(kernel = factor(kernel, levels = kernel_levels))

## Shared aesthetics
kernel_cols <- setNames(c("#1b6ca8", "#e07b39", "#2e8b57"), kernel_levels)
kernel_lty  <- setNames(c("solid", "dashed", "dotdash"),     kernel_levels)
 
base_theme <- theme_bw(base_size = 10) +
  theme(legend.title    = element_blank(),
        legend.position = "bottom",
        panel.grid.minor = element_line(linewidth = 0.15),
        plot.title      = element_text(size = 10, face = "plain"))
 
log_breaks <- 10^seq(-6, 6, by = 2)

## Normalised density, log-log
p_density <- ggplot(kern, aes(t, density, colour = kernel, linetype = kernel)) +
  geom_line(linewidth = 0.6) +
  scale_x_log10(breaks = log_breaks,
                labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))  +
  scale_colour_manual(values = kernel_cols) +
  scale_linetype_manual(values = kernel_lty) +
  labs(title = "(a) normalised kernel",
       x = "time since triggering event (days)",
       y = expression(f[k](t))) +
  base_theme
 
## Cumulative fraction, log time axis
p_cumulative <- ggplot(kern, aes(t, cumulative, colour = kernel, linetype = kernel)) +
  geom_line(linewidth = 0.6) +
  scale_x_log10(breaks = log_breaks,
                labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.02))) +
  scale_colour_manual(values = kernel_cols) +
  scale_linetype_manual(values = kernel_lty) +
  labs(title = "(b) cumulative fraction of triggering",
       x = "time since triggering event (days)",
       y = expression(F[k](t))) +
  base_theme
 
p_density
p_cumulative
```

```{r}
B_ill  <- 0.99
ta_ill <- 100
c_eff  <- (1 - B_ill) * ta_ill / B_ill      

t_grid2 <- 10^seq(-2, 3.7, length.out = 2000)

hybrid <- bind_rows(
  tibble(t = t_grid2, curve = "Rate-state",
         y = g_rs(t_grid2, B_ill, ta_ill)),
  tibble(t = t_grid2, curve = "Omori--Utsu limit (p = 1)",
         y = (1 + t_grid2 / c_eff)^(-1)),
  tibble(t = t_grid2, curve = "Exponential limit",
         y = (1 - B_ill) * exp(-t_grid2 / ta_ill))
) |>
  mutate(curve = factor(curve, levels = c("Rate-state",
                                          "Omori--Utsu limit (p = 1)",
                                          "Exponential limit")))

hy_cols <- c("Rate-state" = "#2e8b57",
             "Omori--Utsu limit (p = 1)" = "grey40",
             "Exponential limit" = "grey40")
hy_lty  <- c("Rate-state" = "solid",
             "Omori--Utsu limit (p = 1)" = "dashed",
             "Exponential limit" = "dotted")

hy_base <- list(
  geom_vline(xintercept = ta_ill, colour = "grey70", linewidth = 0.3),
  annotate("text", x = ta_ill, y = 1, label = "italic(t[a])", parse = TRUE,
           hjust = -0.3, vjust = 1, size = 3, colour = "grey40"),
  geom_line(linewidth = 0.6),
  scale_colour_manual(values = hy_cols),
  scale_linetype_manual(values = hy_lty),
  base_theme
)

## log-log
ggplot(hybrid, aes(t, y, colour = curve, linetype = curve)) +
  hy_base +
  scale_x_log10() + scale_y_log10() +
  coord_cartesian(ylim = c(1e-6, 1.5)) +
  labs(title = "(a) log--log: power-law regime", x = "time (days)",
       y = expression(g[3](t)))

## log-linear
ggplot(hybrid, aes(t, y, colour = curve, linetype = curve)) +
  hy_base +
  scale_y_log10() +
  coord_cartesian(xlim = c(0, 600), ylim = c(1e-6, 1.5)) +
  labs(title = "(b) log--linear: exponential regime", x = "time (days)",
       y = expression(g[3](t)))
```


# References

::: {#refs}
:::
